Adaptative Computation:\\ An Essay in Memory of John H. Holland

### 2015-11-04 paper.tex
a wimpy conclusion

In this short essay, we consider this larger framework, sketching the
recurring themes that were central to Holland's theory of
adaptive systems: (1) discovery and 
%evolutionary 
dynamics of building blocks; (2) internal models and [prediction?] perpetual
novelty; (3) the art [SF doesn't like the term ART] of scientific modeling; and (4) common
properties of complex adaptive systems.  
%We discuss the role these
%themes have played in computer science, and highlight especially the
%ideas that we think remain relevant to today's research agendas.  [MM
%says: Are we going to do this?]

@@

  To succeed, such a search requires first that
  useful building blocks be preserved once discovered, and second that
  there be an efficient mechanism for combining promising building
  blocks into higher-order building blocks.  Holland proposed using an
  analog of Darwinian selection for the first and the idea of crossing
  over in genetics for the second.  Holland referred to the building
  blocks as \emph{schemas.} 
%which correspond mathematically to hyperplanes in the search space. MM: this makes sense only in bit-string example.

@@

\section{The Art of Scientific Modeling}

@@

A given agent-based model encodes a theory about the mechanisms that
are relevant for producing the behavior of interest.  
%Similar to expert systems [MM says: ?]
%Such models are especially useful for studying systems
%that do not have analytic mathematical descriptions.  [MM took out to shorten this paragraph]
%Agent-based models can facilitate interdisciplinary collaborations because the
%underlying rules can be easily communicated.  
[SF: THINKS this next text doesn't make sense until after complex
systems seciton.]
The agent-based models
championed by Holland were typically idealized versions of complex
systems and not intended to provide detailed, domain-specific
predictions.  Instead they were meant to explore possible general
mechanisms of complex systems and thus provide insights that might
lead to more specific, detailed models.  Such idealized models are
akin to what Dennett has called ``intuition pumps''
\cite{Dennett1984}.

@@

\section{Complexity}

@@

Holland was interested in a broad array of adaptive systems---immune
systems, ecologies, financial markets, cities, and the brain---systems
that are \emph{complex}.  In the 1980s, he teamed up with a small
group of scientists, primarily physicists with a sprinkling of
economists and biologists, to discuss what properties this wide swath of systems
have in common.  The discussions helped define the intellectual mission
of the Santa Fe Institute (SFI), the first institution dedicated to
developing a science of complexity.  [add in how many complexity
  institutes there are worldwide and which others John played a hand
  in. MM says: Do we really need that?]  Holland brought to these discussions his lifelong study of
adaptation and a reminder that true theories about complexity would
need to look deeper than phenomenological descriptions but also
account for the `how' and `why' of these systems.
%describing phenomena like chaos, power laws, and ONE MORE, and acoount for the

As the discussions matured, a consensus developed about the basic
elements of complexity: (1) complex systems are composed of many
components with nonlinear interactions; (2) such systems
are characterized by complex \emph{emergent} behavior, exhibiting
higher-order patterns; (3) such systems operate at multiple (and often
nested) spatial and temporal scales, with some behavior being
conserved across all scales and other behaviors changing at different
scales, and (4) these systems exhibit continual adaptation, adjusting
their behavioral rules through evolution and learning.  Although this
is far from a formal definition of complex systems, most people
working in the field today are interested in systems that have these
properties.

To illustrate his ideas about the ubiquity of adaptation, Holland and
co-workers developed two exploratory models, the SFI Artificial Stock
Market and ECHO.

In the early 1990s, Holland teamed up with other Santa Fe Institute researchers,
including several economists, to tackle the mismatch between predictions
of rational expectations theory (the then dominant theory in economics) 
and empirically observed stock market behaviors 
%that deviated from the predicted equilibrium solutions.  
In brief, most
economic theory of the day assumed that all participants in an economy
or financial market are 100\% rational and act to maximize their
individual gain.  In real life, however, actors in economies
and markets are rarely wholly rational, and financial markets often deviate
from rationality, for example with speculative bubbles and crashes.

The SFI Artificial Stock Market project \cite{Arthur1997,Palmer1994} explored a model in which rational
traders are replaced by \emph{adaptive} traders---those who learn to
forecast stock prices over time.  The model allowed testing for the
possible emergence of different trading strategies, including
\emph{fundamental}, \emph{technical}, or \emph{uninformed}
strategies. The simulated market with adaptive trading agents was run
many times, and the dynamics of price and trading volumes were
compared to observed patterns in real markets.  Holland and his
collaborators found that the model's dynamics replicated several
features of real-life markets.

Although the SFI Stock Market model was highly simplified, it was very
influential and led to many follow-on projects.  It was a clear
demonstration of the essential role that adaptation plays in complex
systems, and illustrated how Holland's theories of continual learning
in response to intermittent feedbacks from the environment could be
integrated into domain-specific settings.

ECHO \cite{Holland1999,Hraber1997} was an even more ambitious model
that Holland and collaborators developed during the 1990s.  ECHO's
scope included the evolution of interacting populations of individual
agents, which through competition and learning discover symbiotic
triads such as the famous ant-fly-caterpillar interaction, the Wiksell
triangle of trading relationships in economics, and the immune
system's learned ability to distinguish `self' from `other.'

ECHO formalized Holland's theories about complex adaptive systems into
a runnable computational system where agents evolved external markers
(called tags) and internal preferences, and used them to form higher
level aggregate structures (trading relationships, symbiotic groups,
trophic cascades, interdependent organizations, etc.).  ECHO agents
sometimes discovered mimicry to deceive competitors, and over time,
the model developed increasingly complex structures and behaviors,
often reproducing patterns observed in nature.  For example, ECHO
evolved a diversity of agent types whose rank-frequency distribution
closely parallels the well-known Preston curve in ecology, a
quantitative statement of the adage that ``most species are rare''. Many
of the insights behind this project are described in Holland's book \emph{Hidden Order} \cite{Holland1995}. 
The broad scope of this project was appealing to immunologists, economists, and evolutionary biologists alike.

@@

\section{Relevance to Modern CS}
[MM says:  Do we need this section?  Or can we incorporate these points within the previous sections?]

\begin{itemize}
\item Evolutionary computation
\item Q Learning
\item Backprop
\item    From Turing nomination: "Holland's machine learning system known as the
   Learning Classifier System (LCS), developed in the early 1980s,
   incorporated a reinforcement learning algorithm known as the bucket
   brigade for non-Markovian environments, anticipating by nearly a
   decade non-Markovian learning algorithms."
\item Exploitation versus exploration -- relevance to reinforcement
  learning, other parts of machine learning and optimization.
\item Active learning.  (Two-armed bandit problem.) 
\item On-line learning.  
\end{itemize}

@@

Introduced a few generations of students to computation in natural systems, an idea that today is better accepted.  His insights were deeper and more general than what often passes for work in biomimicry, e.g., for robots.

The ideas have had huge impact and should still be a beacon for research in intelligent and complex systems

John's personality and humanity is inextricably tangled up with his intellectual contributions.

>>>>

In this short essay, we consider this larger framework, sketching the
recurring themes that were central to Holland's theory of
adaptive systems: (1) discovery and 
%evolutionary 
dynamics of building blocks; (2) internal models and prediction; (3) 
exploratory modeling; and (4) complex adaptive systems.  
%We discuss the role these
%themes have played in computer science, and highlight especially the
%ideas that we think remain relevant to today's research agendas.  [MM
%says: Are we going to do this?]

@@

  To succeed, such a search requires first that
  useful building blocks be preserved once discovered, and second that
  there be an efficient mechanism for combining promising building
  blocks into higher-order building blocks.  Holland proposed using an
  analog of Darwinian selection for the first and the idea of crossing
  over in genetics for the second.  Holland referred to the building
  blocks as \emph{schemas} and defined them as hyperplanes in the search space.

@@

\section{Exploratory Modeling}

@@

A given agent-based model encodes a theory about the mechanisms that
are relevant for producing the behavior of interest.  
%Similar to expert systems [MM says: ?]
%Such models are especially useful for studying systems
%that do not have analytic mathematical descriptions.  [MM took out to shorten this paragraph]
%Agent-based models can facilitate interdisciplinary collaborations because the
%underlying rules can be easily communicated.  
%[SF: THINKS this next text doesn't make sense until after complex
%systems seciton.]
The exploratory models
championed by Holland were idealized and not intended to provide detailed, domain-specific
predictions.  Instead they were meant to explore possible general
mechanisms of complex systems and thus provide insights that might
lead to more specific, detailed models.  Such idealized models are
akin to what Dennett has called ``intuition pumps''
\cite{Dennett1984}.

@@

\section{Complex Adaptive Systems}

@@

Holland was interested in a broad array of adaptive systems---immune
systems, ecologies, financial markets, cities, and the brain---systems
that are \emph{complex}.  In the 1980s, he teamed up with a small
group of scientists, primarily physicists with a sprinkling of
economists and biologists, to discuss what properties this wide swath of systems
have in common.  The discussions helped define the intellectual mission
of the Santa Fe Institute (SFI), the first institution dedicated to
developing a science of complexity, and the many others that followed.
% [add in how many complexity
%  institutes there are worldwide and which others John played a hand
%  in. MM says: Do we really need that?]  
Holland brought to these discussions his lifelong study of
adaptation and a reminder that serious theories about complexity would
need to look deeper than phenomenological descriptions but also
account for the `how' and `why' of these systems.
%describing phenomena like chaos, power laws, and ONE MORE, and acoount for the

As the discussions matured, a consensus developed about the basic
elements of complex systems, which are: (1) 
%complex systems are 
composed of many
components with nonlinear interactions; (2) 
%such systems
%are 
characterized by complex \emph{emergent} behavior, exhibiting
higher-order patterns; (3) 
%such systems 
operating at multiple (and often
% interacting? 
nested) spatial and temporal scales, with some behavior being
conserved across all scales and other behaviors changing at different
scales, and (4) 
%these systems exhibit 
adaptative, with 
behavioral rules continually adjusting through evolution and learning.  Although this
is far from a formal definition of complex systems, most people
working in the field today are interested in systems that have these
properties.

%To illustrate his ideas about the ubiquity of adaptation, Holland and
%co-workers developed two exploratory models, the SFI Artificial Stock
%Market and ECHO.

In the early 1990s, Holland teamed up with other Santa Fe Institute researchers,
including several economists, to tackle the mismatch between predictions
of rational expectations theory (the then dominant theory in economics) 
and empirically observed stock market behaviors. 
%that deviated from the predicted equilibrium solutions.  
In brief, most
economic theory of the day assumed that all participants in an economy
or financial market are 100\% rational and act to maximize their
individual gain.  In real life, however, actors in economies
and markets are rarely wholly rational, and financial markets often deviate
from rationality, for example with speculative bubbles and crashes.

The SFI Artificial Stock Market project \cite{Arthur1997,Palmer1994}
developed an exploratory model in which rational
traders were replaced by \emph{adaptive} traders---those who learn to
forecast stock prices over time.  The model tested for the
emergence of different trading strategies, including
\emph{fundamental}, \emph{technical}, or \emph{uninformed}
strategies. The simulated market with adaptive trading agents was run
many times, and the dynamics of price and trading volumes were
compared to observed patterns in real markets.  Holland and his
collaborators found that the model's dynamics replicated several
features of real-life markets.

Although the SFI Stock Market model was highly simplified, it was very
influential and led to many follow-on projects.  It 
demonstrated clearly the essential role that adaptation plays in complex
systems, and it illustrated how Holland's theories of continual learning
in response to intermittent feedbacks from the environment could be
integrated into domain-specific settings.

ECHO \cite{Holland1999,Hraber1997} was an even more ambitious model
that Holland and collaborators developed during the 1990s. 
% ECHO's
% scope included the evolution of interacting populations of individual
% agents, which through competition and learning discover symbiotic
% triads such as the famous ant-fly-caterpillar interaction, the Wiksell
% triangle of trading relationships in economics, and the immune
% system's learned ability to distinguish `self' from `other.'
ECHO formalized Holland's idealization of complex adaptive systems into
a runnable computational system where agents evolved external markers
(called tags) and internal preferences, and then used them to acquire
resources, form higher
level aggregate structures (such trading relationships, symbiotic groups,
trophic cascades, interdependent organizations, etc.).  ECHO agents
sometimes discovered mimicry to deceive competitors, and over time,
the model developed increasingly complex structures and behaviors, at
least up to a point.  ECHO reproduced several well-known patterns 
observed in nature, for example, over time the population of ECHO
evolved agents exhibited diversity of agent types, diversity with a 
rank-frequency distribution that
closely paralleled the well-known Preston curve in ecology.  In
layman's terms, ECHO evolved populations where 
%quantitative statement of the adage that
``most species are rare.''  
The broad scope of the model, together with its ability to produce
easily identifiable patterns, was appealing to immunologists, economists, and evolutionary biologists alike.   Many
of the insights behind this project are described in Holland's third book \emph{Hidden Order} \cite{Holland1995}.

@@

%\section{Relevance to Modern CS}
%[MM says:  Do we need this section?  Or can we incorporate these points within the previous sections?]

@@

Holland was unusual in his ability to absorb the essence of other
disciplines, articulate grand overarching
theories, and them back them up with computational mechanisms and
mathematics.  Unlike most researchers he moved seamlessly among these
three modes of thinking, developing models that were years ahead of their time.  A
close reading of his work reveals the antecedents of many ideas prevalent in
machine learning today, e.g., reinforcement, learning in non-Markovian
environments, active learning, etc.   His seminal genetic algorithm
spawned the field of evolutionary computation, and his insights
and wisdom helped define what are now referred to as the
\emph{sciences of complexity}.   With such a broad scope, it is easy
to pick at the details.

His many books and articles have influenced 
scientists around the work and across many disciplines.  Closer to home, he introduced several
generations of students at the University of Michigan to 
computation in natural systems, an idea that even today
remains somewhat controversial, despite successes in biomimicry
for robotics, abstractions of phermone trails in ant colonies for
optimization methods, the use of mechanisms from immunology to
improve computer security, and recent work on technological
evolution.

Behind the ideas, of course, is the man himself.   Everyone who knew
John personally will remember the gleam in his eye when he
encountered a new idea; his willingness to talk to anyone, no matter
how famous or how unfamous; and his incredible generosity and patience.
%His insights aboutwere deeper and more general than what often passes for work in biomimicry, e.g., for robots.
Holland's personality and humanity were somehow inextricably tangled
up with his intellectual contributions, and although his ideas will serve as a beacon
for research in intelligent and complex systems
for many years to come, the world of science is poorer without John
Holland on the scene.

% \begin{itemize}
% \item Evolutionary computation
% \item Q Learning
% \item Backprop
% \item    From Turing nomination: "Holland's machine learning system known as the
%    Learning Classifier System (LCS), developed in the early 1980s,
%    incorporated a reinforcement learning algorithm known as the bucket
%    brigade for non-Markovian environments, anticipating by nearly a
%    decade non-Markovian learning algorithms."
% \item Exploitation versus exploration -- relevance to reinforcement
%   learning, other parts of machine learning and optimization.
% \item Active learning.  (Two-armed bandit problem.) 
% \item On-line learning.  
% \end{itemize}
